---
title: "Lab 7 Homework"
author: "Liana Williams"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
```

For this assignment we are going to work with a large data set from the [United Nations Food and Agriculture Organization](http://www.fao.org/about/en/) on world fisheries. These data are pretty wild, so we need to do some cleaning. First, load the data.  


Load the data `FAO_1950to2012_111914.csv` as a new object titled `fisheries`.
```{r}
fisheries <- readr::read_csv(file = "data/FAO_1950to2012_111914.csv")
```

1. Do an exploratory analysis of the data (your choice). What are the names of the variables, what are the dimensions, are there any NA's, what are the classes of the variables? 

This gives an overall summary! 
```{r}
#summary(fisheries)
```

Names of the variables. 
```{r}
names(fisheries) 
```

Dimensions of the dataset
```{r}
dim(fisheries) #There are 17692 observations(rows), 71 variables(columns)
```

Determine if there is any NAs. 
```{r}
anyNA(fisheries) #The output of the function is "true" so, there are NAs in the file
```

Outputs the data classes of the variables 
```{r}
glimpse(fisheries)
```

#Alternative answer
```{r}
#glimpse(fisheries)
```

```{r}
#head(fisheries)
```

```{r}
#tail(fisheries)
```

```{r}
#str(fisheries)
```
2. Use `janitor` to rename the columns and make them easier to use. As part of this cleaning step, change `country`, `isscaap_group_number`, `asfis_species_number`, and `fao_major_fishing_area` to data class factor. 
```{r}
fisheries <- clean_names(fisheries) #clean the data. 
```

```{r}
names(fisheries) #check that the variables changed
```

```{r}
fisheries$country <- as.factor(fisheries$country)
fisheries$isscaap_group_number <- as.factor(fisheries$isscaap_group_number)
fisheries$asfis_species_number <- as.factor(fisheries$asfis_species_number)
fisheries$fao_major_fishing_area <- as.factor(fisheries$fao_major_fishing_area)
#The $ allows access to one variable in a data set.
```

#Alternative to Q#2
```{r}
#fisheries <- janitor::clean_names(fisheries)
#names(fisheries)
```

```{r}
#fisheries$country <- as.factor(fisheries$country)
#fisheries$isscaap_group_number <- as.factor(fisheries$isscaap_group_number)
#fisheries$asfis_species_number <- as.factor(fisheries$asfis_species_number)
#fisheries$fao_major_fishing_area <- as.factor(fisheries$fao_major_fishing_area)
```

Here is a nice alternate way to change classes using dplyr.  
```{r}
#fisheries <- 
  #fisheries %>% mutate(across(c(country, isscaap_group_number, asfis_species_number, fao_major_fishing_area), as_factor))
```
 
We need to deal with the years because they are being treated as characters and start with an X. We also have the problem that the column names that are years actually represent data. We haven't discussed tidy data yet, so here is some help. You should run this ugly chunk to transform the data for the rest of the homework. It will only work if you have used janitor to rename the variables in question 2!  
```{r}
fisheries_tidy <- fisheries %>% 
  pivot_longer(-c(country,common_name,isscaap_group_number,isscaap_taxonomic_group,asfis_species_number,asfis_species_name,fao_major_fishing_area,measure),
               names_to = "year",
               values_to = "catch",
               values_drop_na = TRUE) %>% 
  mutate(year= as.numeric(str_replace(year, 'x', ''))) %>% 
  mutate(catch= str_replace(catch, c(' F'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('...'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('-'), replacement = '')) %>% 
  mutate(catch= str_replace(catch, c('0 0'), replacement = ''))

fisheries_tidy$catch <- as.numeric(fisheries_tidy$catch)
```

**This was added to clean the data, when doing the homework remove all the hashtags and run this**

3. How many countries are represented in the data? Provide a count and list their names.
```{r}
fisheries_tidy %>% 
  summarize(n_countries = n_distinct(country))
```

**There are 203 countries** 

```{r}
fisheries_tidy %>% 
  count(country) #This is a count for the list of country names
```

#Alternative to Q3 
```{r}
#fisheries_tidy %>% 
  #summarize(ncountries = n_distinct(country))
```

```{r}
#levels(fisheries_tidy$country)
```

4. Refocus the data only to include country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch.
```{r}
fisheries_new <- fisheries_tidy %>% 
  select(country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch)
fisheries_new
```

#Alternative to Q4 
```{r}
#fisheries_tidy_focused <- fisheries_tidy %>% 
  #select(country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch)
#fisheries_tidy_focused
```
5. Based on the asfis_species_number, how many distinct fish species were caught as part of these data?
```{r}
fisheries_new %>%
  summarize(distinct_fish_species = n_distinct(asfis_species_number))
```
**1551 species were caught as part of the data**

#Alternative to Q5 
```{r}
#fisheries_tidy_focused %>% 
  #summarize(n_taxa=n_distinct(asfis_species_number))
```

6. Which country had the largest overall catch in the year 2000?
```{r}
fisheries_new %>% 
  filter(year == 2000) %>% #filter by year 2000
  group_by(country) %>% #show country (categorical)  
  summarize(overall_catch = sum(catch, na.rm = TRUE))%>% #Calc. overall catch value  
  arrange(desc(overall_catch)) #sort data to find largest overall catch
  
```
**The country with the largest overall catch in the year 2000 was China. 25,899 fish were caught!**

#Alternative to Q6 
```{r}
#fisheries_tidy_focused %>% 
  #filter(year==2000) %>%
  #group_by(country) %>% 
  #summarize(catch_total=sum(catch, na.rm=T)) %>% 
  #arrange(desc(catch_total))
```

7. Which country caught the most sardines (_Sardina pilchardus_) between the years 1990-2000?
```{r}
fisheries_new %>% 
  filter(year >= 1990, year <= 2000, asfis_species_name == "Sardina pilchardus") %>% #filter time and sardines 
  group_by(country) %>% 
  summarize(total_sardines = sum(catch, na.rm = T)) %>% #within 1990-2000 it shows #'s sardines caught in each country
  arrange(desc(total_sardines)) #show which countries caust the most sardines 
```

**The country that caught the most sardines between 1990-2000 was Morocco with 7,470 sardines.**

#Alternative to Q7 
```{r}
#fisheries_tidy_focused %>% 
  #filter(asfis_species_name== "Sardina pilchardus") %>% 
  #filter(year>=1990 & year<=2000) %>% 
  #group_by(country) %>% 
  #summarize(catch_total=sum(catch, na.rm=T)) %>% 
  #arrange(desc(catch_total))
```

Double check
```{r}
#sardines1 <- fisheries_tidy_focused %>% 
  #filter(stringr::str_detect(asfis_species_name, "Sardina")) %>% 
  #filter(year>=1990 & year<=2000)
```

```{r}
#sardines1 %>% 
  #group_by(country) %>% 
  #summarize(catch_total=sum(catch, na.rm=T)) %>% 
  #arrange(desc(catch_total))
```

Triple check
```{r}
#sardines2 <- fisheries_tidy_focused %>% 
  #filter(stringr::str_detect(isscaap_taxonomic_group, "sardines")) %>% 
  #filter(year>=1990 & year<=2000)
```

```{r}
#sardines2 %>% 
  #group_by(country) %>% 
  #summarize(catch_total=sum(catch, na.rm=T)) %>% 
  #arrange(desc(catch_total))
```

8. Which five countries caught the most cephalopods between 2008-2012?
```{r}
fisheries_new %>% 
  filter(year >= "2008", year <= "2012", asfis_species_name == "Cephalopoda") %>% 
  group_by(country) %>% 
  summarize(total_cephalopods = sum(catch, na.rm = T)) %>% 
  arrange(desc(total_cephalopods)) %>% #countries with most cephalopods caught is at the top
  head()
```
**Five countries that caught the most cephalopods between 2008-2012 is India, China, Spain, Algeria, and France in that order**

#Alternative to Q8 
```{r}
#fisheries_tidy_focused %>% 
  #filter(str_detect(isscaap_taxonomic_group, "Squids")) %>% 
  #filter(year>=2008 & year<=2012) %>% 
  #group_by(country) %>% 
  #summarize(catch_total=sum(catch, na.rm=T)) %>% 
  #arrange(desc(catch_total))
```

```{r}
#fisheries_tidy_focused %>% 
  #filter(isscaap_taxonomic_group=="Squids, cuttlefishes, octopuses") %>% 
  #count(asfis_species_name, sort=T)
```

9. Which species had the highest catch total between 2008-2012? (hint: Osteichthyes is not a species)
```{r}
fisheries_new %>% 
  filter(year >= "2008", year <= "2012") %>% 
  group_by(asfis_species_name) %>% 
  summarize(total_catch = sum(catch, na.rm = T)) %>% #find the total number of fish caught btwn 2008-2012
  arrange(desc(total_catch)) #the species that was caught the most is at the top
  
```

#Alternative to Q9
```{r}
#fisheries_tidy_focused %>% 
  #filter(year>=2008 & year<=2012)%>% 
  #group_by(asfis_species_name) %>% 
  #summarize(catch_total=sum(catch, na.rm=T)) %>% 
  #arrange(desc(catch_total))
```

10. Use the data to do at least one analysis of your choice.
```{r}
fisheries_new %>% 
  group_by(year) %>% 
  summarize(mean_catch = mean(catch, na.rm = T), min_catch = min(catch, na.rm = T), max_catch = max(catch, na.rm = T))
#This analysis shows the mean, min, and max amounts of fish caught during each year from 1950-2012
```

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.   