---
title: "Homework 7"
author: "Liana Williams"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(naniar)
``` 

## Data


**1. For this homework, we will use two different data sets. Please load `amniota` and `amphibio`.**  

`amniota` data:  
Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history
database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109.
doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r}
getwd()
```


```{r}
amniota <- read_csv("/Users/liana/Desktop/BIS15/BIS15W2024_lwilliams/lab8/data/amniota.csv") %>% clean_names()
```

`amphibio` data:  
Oliveira BF, São-Pedro VA, Santos-Barrera G, Penone C, Costa GC (2017). “AmphiBIO, a global database
for amphibian ecological traits.” _Scientific Data_, *4*, 170123. doi: 10.1038/sdata.2017.123 (URL:
https://doi.org/10.1038/sdata.2017.123).
```{r}
amphibio <- read_csv("/Users/liana/Desktop/BIS15/BIS15W2024_lwilliams/lab8/data/amphibio.csv") %>% clean_names()
```

## Questions  
**2. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  
```{r}
glimpse(amniota)
```
NA's in the dataset have a value of -999.

```{r}
miss_var_summary(amniota) #total NA's in this dataset
```

**3. Do some exploratory analysis of the `amphibio` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  
```{r}
glimpse(amphibio)
```

NA's are listed as NA. 

```{r}
summary(amphibio) #variables dir,lar,viv include 0s which we learned in class that sometimes that is used for NAs
```

```{r}
miss_var_summary(amphibio) #amount of NA's in this dataset
```

**4. How many total NA's are in each data set? Do these values make sense? Are NA's represented by values?**   
```{r}
amniota %>% 
  map_df(~ sum(is.na(.)))
```

```{r}
summary(amniota)
```

No, NAs in the amniota dataset. This does not make sense. When using the summary function I notice values such as (-999, -999.0, and -999.000).

```{r}
amphibio %>% 
  map_df(~ sum(is.na(.)))
```

```{r}
summary(amphibio)
```

Yes, NAs in the amphibio dataset, 170566 total. It doesn't make sense since there are many NAs in the columns.The summary function makes it seem like there are no NA's represented by values.

**5. Make any necessary replacements in the data such that all NA's appear as "NA".**   
```{r}
amniota <- amniota %>% 
  replace_with_na_all(condition = ~.x == -999)
```

All the values the end with body_mass_g have 0 for their minimum value which isn't biologically accurate. 

```{r}
amniota %>% 
  select(female_body_mass_g, male_body_mass_g, no_sex_body_mass_g, adult_body_mass_g) %>% 
  summarize(across(contains("body_mass"), min, na.rm = T))
```

Amphibio: 
```{r}
amphibio <- amphibio %>% 
  replace_with_na_all(condition = ~.x == 0)
amphibio
```

```{r}
summary(amphibio)
```

**6. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amniota` data.**  
```{r}
naniar::miss_var_summary(amniota)
```

**7. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amphibio` data.**
```{r}
naniar::miss_var_summary(amphibio)
```

**8. For the `amniota` data, calculate the number of NAs in the `egg_mass_g` column sorted by taxonomic class; i.e. how many NA's are present in the `egg_mass_g` column in birds, mammals, and reptiles? Does this results make sense biologically? How do these results affect your interpretation of NA's?**  
```{r}
amniota %>% 
  select(egg_mass_g, class) %>% 
  group_by(class) %>% 
  miss_var_summary()
```
NAs exist in each of the classes with Aves (4914), Mammalia (4953), and Reptilita(6040). This result makes sense biologically since most mammals do not lay eggs so, for the rare amount of mammals that do lay eggs it was probably not recorded thats why percent missing is 100. The results of this code chunk change my interpretation of the NA's since there is biologically to account for the missing data and not just the fact that it wasn't measured. 

**9. The `amphibio` data have variables that classify species as fossorial (burrowing), terrestrial, aquatic, or arboreal.Calculate the number of NA's in each of these variables. Do you think that the authors intend us to think that there are NA's in these columns or could they represent something else? Explain.**
```{r}
amphibio %>% 
  select(fos, ter, aqu, arb) %>%  #classify the species
  naniar::miss_var_summary()
amphibio
```
I think the author's don't intend for us to think there are NAs since they trying to classify the species based on whether they fit (1 = true) into the four variables fos, arb, aqu, and ter. If the species doesn't fit into these four then, it is potentially given the NA value (false).

**10. Now that we know how NA's are represented in the `amniota` data, how would you load the data such that the values which represent NA's are automatically converted?**
```{r}
amniota <- read_csv("/Users/liana/Desktop/BIS15/BIS15W2024_lwilliams/lab8/data/amniota.csv", na = c("NA", " ", ".", "-999")) %>% clean_names()
```

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.  