---
title: "Homework 10"
author: "Liana Williams"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(naniar)
``` 

## Desert Ecology
For this assignment, we are going to use a modified data set on [desert ecology](http://esapubs.org/archive/ecol/E090/118/). The data are from: S. K. Morgan Ernest, Thomas J. Valone, and James H. Brown. 2009. Long-term monitoring and experimental manipulation of a Chihuahuan Desert ecosystem near Portal, Arizona, USA. Ecology 90:1708.
```{r}
deserts <- read_csv("data/surveys_complete.csv") %>% clean_names()
```

1. Use the function(s) of your choice to get an idea of its structure, including how NA's are treated. Are the data tidy?  
```{r}
glimpse(deserts) #structure of the data
```

```{r}
miss_var_summary(deserts) #The NA's are represented by NA
```

```{r}
summary(deserts)
```

```{r}
deserts
#the data is tidy! 
# each variable has its own column  
# each observation has its own row  
# each value has its own cell  
```



2. How many genera and species are represented in the data? What are the total number of observations? Which species is most/ least frequently sampled in the study?
```{r}
deserts %>% 
  summarize(n_genera = n_distinct(genus), #amount of genera in the dataset
            n_species = n_distinct(species), #amount of distinct species in dataset
            n_total_observations = n()) #total amount of observations
```
In the dataset: 26 genera, 40 species, 34,786 observations. 
```{r}
deserts %>% 
  count(species) %>% #amount of observations for each species
  arrange(n)
```
^Species that is the least frequently sampled have one observation. (clarki, scutalatus, tereticaudus, tigris, uniparens, viridis)

```{r}
deserts %>% 
  count(species) %>%
  arrange(desc(n)) #sort most to least observations
```
^Species that is the most frequently sampled merriami with 10,596!

3. What is the proportion of taxa included in this study? Show a table and plot that reflects this count.
```{r}
taxa_prop <- deserts %>% 
  tabyl(taxa)
taxa_prop
```

```{r}
taxa_prop %>% 
  ggplot(aes(x = taxa, y = n, fill = taxa)) +
  geom_col() + 
  scale_y_log10() + 
  labs(title = "Taxa Represented in Study",
       x = "Taxa",
       y = "Number of Observations") + 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.75)) #move and resize labels
```
```{r}
taxa_prop %>% 
  ggplot(aes(x = taxa, y = percent, fill = taxa)) + #changing the fill to be based on taxa
  geom_col() + 
  labs(title = "Taxa Represented in Study",
       x = "Taxa",
       y = "Proportion") + #adding labels
  theme(plot.title = element_text(size = rel(1.3), hjust = 0.5)) 
```

4. For the taxa included in the study, use the fill option to show the proportion of individuals sampled by `plot_type.`
```{r}
deserts %>% 
  ggplot(aes(x = taxa, fill = plot_type))+ #this is the modified line to show the prop. of indiv. sampled by 'plot_type'
  geom_bar()+
  scale_y_log10()+ #fixes viewing for continuous data
  labs(title = "Taxa Represented in Study", 
       x = "Taxa", 
       y = "Number of Observations",
       fill = "Type of Plot") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.75))
```

5. What is the range of weight for each species included in the study? Remove any observations of weight that are NA so they do not show up in the plot.
```{r}
deserts %>% 
  filter(weight != "NA") %>% 
  unite(species_name, "genus", "species", sep = " ") %>% 
  ggplot(aes(x = species_name, y = weight))+
  geom_boxplot()+
  coord_flip()+
  labs(title = "Weight by Species",
       x = "Weight",
       y = "Species Name")+
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```


6. Add another layer to your answer from #5 using `geom_point` to get an idea of how many measurements were taken for each species.
```{r}
deserts %>% 
  filter(weight != "NA") %>% 
  unite(species_name, "genus", "species", sep = " ") %>% 
  ggplot(aes(x = species_name, y = weight))+
  geom_boxplot()+
  coord_flip()+
  labs(title = "Weight by Species",
       x = "Weight",
       y = "Species Name")+
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))+
  geom_point(aes(color = weight), size = 0.5) #Modified line of code that adds a scatter plot
```


7. [Dipodomys merriami](https://en.wikipedia.org/wiki/Merriam's_kangaroo_rat) is the most frequently sampled animal in the study. How have the number of observations of this species changed over the years included in the study?
```{r}
deserts %>% 
  filter(species == "merriami") %>%  #filter by most freq animal
  group_by(year) %>% 
  summarize(n_observations = n())
```

```{r}
deserts %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) + 
  geom_point() + 
  labs(title = "Number of Observations by Year",
       x = "Year",
       y = "Number of Observations") + 
  theme(plot.title = element_text(size = rel(1.3), hjust = 0.5)) + 
  geom_smooth(method = lm, se = T) #adding line of best fit and it shows standard error
```
As time has gone on, the number of observations seems to have a positive correlation which could mean the more species were observed over time due to biological factors like predators typically preferring larger prey or the ability for Merriam's Kangaroos to produce up to three litters per year. 

8. What is the relationship between `weight` and `hindfoot` length? Consider whether or not over plotting is an issue.
```{r}
deserts %>% 
  ggplot(aes(x = weight, y = hindfoot_length)) + 
  geom_jitter() + #reducing overplotting
  labs(title = "Hindfoot Length vs Weight",
       x = "Weight",
       y = "Hindfoot Length") + #adding labels
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5)) + 
  geom_smooth(method = lm, se = T)
```
There is a positive correlation between weight and hindfoot length. Overplotting is an issue. 
#DOUBLE CHECK ^^^^^^^

9. Which two species have, on average, the highest weight? Once you have identified them, make a new column that is a ratio of `weight` to `hindfoot_length`. Make a plot that shows the range of this new ratio and fill by sex.
```{r}
deserts %>% 
  group_by(species) %>% 
  summarize(avg_weight = mean(weight, na.rm = T)) %>% 
  arrange(desc(avg_weight)) %>% 
  head(2, species) #displays the two species with highest average weight
```
The two species are albigula and spectabilis!

```{r}
#New column with ratio of 'weight' to 'hind_foot' 
high_weight_desert <- deserts %>% 
  filter(species == "albigula" | species == "spectabilis") %>% 
  mutate(w_hindfoot_length_ratio = weight / hindfoot_length)
high_weight_desert 
```
#^^^^^^DOUBLE CHECK THIS

```{r}
high_weight_desert %>% 
  filter(sex != "NA") %>% 
  ggplot(aes(x = species, y = w_hindfoot_length_ratio, fill = sex))+
  geom_boxplot(alpha = 0.5)
```


10. Make one plot of your choice! Make sure to include at least two of the aesthetics options you have learned.
#The plot shows the range of weights for male versus female animals comparing the plot type.     
^^^^^^DO SOMETHING ELSE
```{r}
#deserts %>% 
  #filter(sex != "NA") %>% #filtering out the NA's in the sex column
  #ggplot(aes(x = sex, y = weight, fill = plot_type)) + #creating the plot
  #geom_boxplot(alpha = 0.5) + #creating a boxplot
  #labs(title = "Weight by Sex",
      # x = "Sex",
      # y = "Weight",
     #  fill = "Taxa") + #adding labels
 # theme(plot.title = element_text(size = rel(1.3), hjust = 0.5)) + #moving and resizing labels
 # scale_y_log10() #for ease of viewing
```

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 