---
title: "Homework 12"
author: "Liana Williams"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries  
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(ggmap)
``` 

## Load the Data
We will use two separate data sets for this homework.  

1. The first [data set](https://rcweb.dartmouth.edu/~f002d69/workshops/index_rspatial.html) represent sightings of grizzly bears (Ursos arctos) in Alaska.  


2. The second data set is from Brandell, Ellen E (2021), Serological dataset and R code for: Patterns and processes of pathogen exposure in gray wolves across North America, Dryad, [Dataset](https://doi.org/10.5061/dryad.5hqbzkh51).  

1. Load the `grizzly` data and evaluate its structure.  
```{r}
grizzly <- read_csv("data/bear-sightings.csv") %>% clean_names()
```
```{r}
glimpse(grizzly)
```


2. Use the range of the latitude and longitude to build an appropriate bounding box for your map.
```{r}
#Det. Range of Lat. & Lon. 
grizzly %>% 
  select(latitude, longitude) %>% 
  summary()
```
```{r}
#Build Bounding Box for Map 
lat_grizzly <- c(55.02, 70.37)
lon_grizzly <- c(-166.2, -131.3)
bbox_grizzly <- make_bbox(lon_grizzly, lat_grizzly, f = 0.03) 
```

3. Load a map from `stamen` in a terrain style projection and display the map. 
```{r}
register_stadiamaps("ae9b303b-2453-44b9-8f34-55a7a872cf5c", write = FALSE)
```

```{r}
#Load Map
map_grizzly <- get_stadiamap(bbox_grizzly, maptype = "stamen_terrain", zoom = 7)
```

```{r}
#Display Map
ggmap(map_grizzly)
```


4. Build a final map that overlays the recorded observations of grizzly bears in Alaska.
```{r}
#create map, overlay, add labels, adjust
ggmap(map_grizzly)+
  geom_point(data = grizzly, aes(longitude, latitude), size = 0.3)+
  labs(x= "Longitude", 
       y= "Latitude",
       title = "Location of Grizzly Bears in Alaska")+
  theme(plot.title = element_text(size = rel(1.3), hjust = 0.5))
  
```


Let's switch to the wolves data. Brandell, Ellen E (2021), Serological dataset and R code for: Patterns and processes of pathogen exposure in gray wolves across North America, Dryad, [Dataset](https://doi.org/10.5061/dryad.5hqbzkh51).  

5. Load the data and evaluate its structure.  
```{r}
wolves <- read_csv("data/wolves_data/wolves_dataset.csv") %>% clean_names()
```
```{r}
glimpse(wolves)
```



6. How many distinct wolf populations are included in this study? Mae a new object that restricts the data to the wolf populations in the lower 48 US states. 
```{r}
n_distinct(wolves$pop) #17 distinct
```

```{r}
tabyl(wolves$pop)
```

```{r}
#New object with wolf pop data in lower 48 states
low48_wolves <- wolves %>% 
  filter(lat > 25.84, lat < 49.38) %>% 
  filter(long > -124.67, long < -66.95)
```

7. Use the range of the latitude and longitude to build an appropriate bounding box for your map. 
```{r}
#Range of Lat & Lon
low48_wolves %>% 
  select(lat, long) %>% 
  summary()
```
```{r}
#Bounding Box for Map 
lat_wolves <- c(33.89, 47.75)
long_wolves <- c(-110.99, -86.82)
bbox_wolves <- make_bbox(long_wolves, lat_wolves, f = 0.03)
```



8.  Load a map from `stamen` in a `terrain-lines` projection and display the map.  
```{r}
map_wolves <- get_stadiamap(bbox_wolves, maptype = "stamen_terrain_lines", zoom = 7)
```

```{r}
ggmap(map_wolves)
```

9. Build a final map that overlays the recorded observations of wolves in the lower 48 states. 
```{r}
ggmap(map_wolves) + 
  geom_point(data = low48_wolves, aes(long, lat), size = 0.6) +
  labs(x= "Longitude",
       y= "Latitude",
       title="Location of Wolves in Lower 48 States")  
```


10. Use the map from #9 above, but add some aesthetics. Try to `fill` and `color` by population.  
```{r}
ggmap(map_wolves) + 
  geom_point(data = low48_wolves, aes(long, lat, fill = pop, color = pop), size = 1.3) + 
  labs(x= "Longitude",
       y= "Latitude",
       title="Location of Wolves in Lower 48 States") + 
  theme(plot.title = element_text(size = rel(1.3), hjust = 0.5))
```


## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 