---
title: "Midterm 1 W24"
author: "Liana Williams"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code must be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above. 

Your code must knit in order to be considered. If you are stuck and cannot answer a question, then comment out your code and knit the document. You may use your notes, labs, and homework to help you complete this exam. Do not use any other resources- including AI assistance.  

Don't forget to answer any questions that are asked in the prompt!  

Be sure to push your completed midterm to your repository. This exam is worth 30 points.  

## Background
In the data folder, you will find data related to a study on wolf mortality collected by the National Park Service. You should start by reading the `README_NPSwolfdata.pdf` file. This will provide an abstract of the study and an explanation of variables.  

The data are from: Cassidy, Kira et al. (2022). Gray wolf packs and human-caused wolf mortality. [Dryad](https://doi.org/10.5061/dryad.mkkwh713f). 

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the wolves data
In these data, the authors used `NULL` to represent missing values. I am correcting this for you below and using `janitor` to clean the column names.
```{r message=FALSE, warning=FALSE}
wolves <- read.csv("data/NPS_wolfmortalitydata.csv", na = c("NULL")) %>% clean_names()
```

## Questions
Problem 1. (1 point) Let's start with some data exploration. What are the variable (column) names?
```{r}
names(wolves)
```
^The variable names are listed above. 

Problem 2. (1 point) Use the function of your choice to summarize the data and get an idea of its structure.
```{r}
glimpse(wolves) # information dense summary of data 
```

Problem 3. (3 points) Which parks/ reserves are represented in the data? Don't just use the abstract, pull this information from the data.
```{r}
table(wolves$park)
```

The five parks represented in the data are: DENA, GNTP, VNP, YNP, and YUCH. 

Problem 4. (4 points) Which park has the largest number of wolf packs?
```{r}
wolves %>% 
  arrange(desc(packsize_aug))
```

The park with the largest number of wolf packs is YNP with 37. packsize_aug is the pack size in the fall. 

Problem 5. (4 points) Which park has the highest total number of human-caused mortalities `mort_all`?
```{r}
wolves %>% 
  select(park, mort_all) %>% 
  arrange(park)
```
mortALL is the total number of human-caused mortalities for the pack-year. DENA park has the highest number (4) of human-caused mortalities. 

The wolves in [Yellowstone National Park](https://www.nps.gov/yell/learn/nature/wolf-restoration.htm) are an incredible conservation success story. Let's focus our attention on this park.  

Problem 6. (2 points) Create a new object "ynp" that only includes the data from Yellowstone National Park.
```{r}
ynp <- wolves %>%
  filter(park=="YNP")
```
The object ynp includes the data from Yellowstone National Park. 

Problem 7. (3 points) Among the Yellowstone wolf packs, the [Druid Peak Pack](https://www.pbs.org/wnet/nature/in-the-valley-of-the-wolves-the-druid-wolf-pack-story/209/) is one of most famous. What was the average pack size of this pack for the years represented in the data?
```{r}
ynp %>% 
  filter(biolyr>=1997 & biolyr<=2010) %>% 
  filter(pack=="druid") %>% 
  summarize(mean_pack_size = mean(packsize_aug))
```
The average pack size for druids between 1997 to 2010 is 14.57143. 

Problem 8. (4 points) Pack dynamics can be hard to predict- even for strong packs like the Druid Peak pack. At which year did the Druid Peak pack have the largest pack size? What do you think happened in 2010?
```{r}
#select(ynp, packsize_aug) %>% 
#filter(pack=="druid")
  #filter(biolyr>=1997 & biolyr<=2010) %>%
  #filter(pack=="druid")
  #summarize(mean_pack_size = mean(packsize_aug))
#view(ynp) #I viewed by name druid in the packs variable and looked at packsizeAug to determine the size. 
```

The druid had the largest pack size in 2001 with 37. In 2010 there are 0 in the pack and I think they were all killed off by another species. 

Problem 9. (5 points) Among the YNP wolf packs, which one has had the highest overall persistence `persisty1` for the years represented in the data? Look this pack up online and tell me what is unique about its behavior- specifically, what prey animals does this pack specialize on?
```{r}
ynp %>% 
  filter(biolyr>=1995 & biolyr<=2021) %>% 
  group_by(pack) %>%   
  summarize(persistence=sum(persisty1, na.rm=T)) %>% 
  arrange(desc(persistence))
```

The pack with the highest overall persistence was mollies with 26. The Mollies are known for its adaptations to killing bison. 

Problem 10. (3 points) Perform one analysis or exploration of your choice on the `wolves` data. Your answer needs to include at least two lines of code and not be a summary function.
```{r}
wolves %>%
  filter(packsize_aug<3) %>% 
  summarize(mean_packsize = mean(packsize_aug)) 
```
