---
title: "code_MT1"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Keyboard Shortcuts 
Insert Pipe Operator Cmd+Shift+M
Knit Document (knitr)	Ctrl+Shift+K	Cmd+Shift+K

Insert chunk (Sweave and Knitr)	Ctrl+Alt+I	Cmd+Option+I
Run all chunks above the current chunk with Command + Option + P on a Mac; Ctrl + Alt + P on Linux and Windows.

Run the current chunk with Command + Option + C or Command + Shift + Enter on a Mac; Ctrl + Alt + C or Ctrl + Shift + Enter on Linux and Windows.

Run the next chunk with Command + Option + N on a Mac; Ctrl + Alt + N on Linux and Windows.

Run all chunks with Command + Option + R or Command + A + Enter on a Mac; Ctrl + Alt + R or Ctrl + A + Enter on Linux and Windows.


**dplyr** is a package in tidyverse
Note: =assignment but, == equal to 

##Syntax
# %>% Piping
passes object on left hand side as first argument (or . argument) of function on righthand side. 
x %>% f(y) is the same as f(x, y) 
```{r}
iris %>% 
  group_by(species) %>% 
  summarize(avg = mean(Sepal.Width)) %>% 
  arrange(avg)
```

#glimpse()
shows all variables, their first value, type, gives number of variables and observations.; information dense summary of tbl data 
```{r}
glimpse(table2)
```
#view() 
puts the table in the environment panel and you can sort ; views like a spreadsheet 
```{r}
view(table2)
```
#arrange()
allows you to sort 
```{r}
table2 %>% 
  arrange(year) #this could also be arrange(year, type)
Ex: 
  arrange(mtcars, mpg) #order rows by values of a column (low to high)
Ex:
  arrange(mtcars, desc(mpg)) #order rows by values of a column (high to low)
```

#select()
include the column names (variables) that you want or you can put a (-) in front of them indicating that you pull out those columns. 
```{r}
table1 %>% 
  select(country, populations)
table1 %>% 
  select(-cases, -year) #select all columns except cases and year 
Ex:
select(iris, contains(".")) #select column whose name contains a character string. 
Ex: 
select(iris, ends_with("length")) #select column whose name ends with a character string. #starts__with is also a command 
Ex: select(iris, sepal.length:petal.width) #select all columns btwn speal.length and petal.width (inclusive)
```
#filter()
Extract rows (observations) that meet specific criteria
```{r}
table1 %>% 
  filter(year == 2000)
table1 %>% 
  filter(year != 2000)
```
#mutate()
create a new column 
```{r}
table1 %>% 
  mutate(fill_it = "with a smile :)") %>% #fill_it is the name of the new variable
  mutate(popul_per_case = population / cases)

table1 %>% 
  mutate(fill_it = "with a smile :)") %>% 
  rename(much_better_name = fill_it)

#mutate() functions take a vector of values and return another vector of values, such as: 
between() #are the values btwn a and b?
```
#group_by()
groups data into rows of the same value of species 
#summarize()
summarize data into a single row of values for each of the groups. 
```{r}
table1 %>% 
  summarize(counting = n())
table1 %>% 
  summarize(sum = sum(population)) #sums all the values in the column 
table1 %>% 
  summarize(average = mean(population))
table1 %>% 
  summarize(avr_popul = mean(population), all_cases = sum(cases), how_many = n())
table1 %>% 
  group_by(cyl, am) %>% 
  summarize(avr_mpg = mean(mpg), avr_hp = mean(hp), count = n())
```
#Distinct() 
pass whatever columns you want and gives distinct rows; remove duplicate rows 
```{r}
distinct(iris)
```

dplyr::sample_n (iris, 10, replace = TRUE) #randomly select n rows
#Count()
counts the number of rows with each unique value of variable



**summary functions** this is wha 'summarize()' uses 
Take a vector of values and return a single value, such as: 
dplyr::first #first value of a vector 
dplyr::last #last value of a vector 
dplyr::nth #nth value of a vector 
dplyr::n #number of values in a vector 
dplyr::n_distinct # number of distinct values in a vector 

IQR 
min #min value in a vector 
max 
mean 
median 
var #variance of a vector 
sd #standard deviation from a vector 

#Logic in R 
<
>
==
<=
>=
!=
%in%
is.na
!is.na 
&, |, !, xor, any, all (Boolean Operators)


AND	&	#conditions on both sides need to be satisfied. 
4 == 3 & 1 == 1

4 == 4 & 1 == 1

FALSE

TRUE

OR	|	#requires at least one condition be met on either side of the boolean operator. Note: the output will be TRUE if both conditions are met
4 == 4 | 1 == 1

4 == 3 | 1 == 1

4 == 3 | 1 == 2

TRUE

TRUE

FALSE

NOT	!	#reverse the outcome of a condition
!(4 == 3) #!(false) = true

!(4 == 4)

TRUE

FALSE

```{r}
library(dplyr)
library(readr)
billboard100 <- read_csv("billboard100.csv")
```

# Piping %>%
```{r}
head(billboard100, 10)

billboard100 %>% head(10)
10 %>% head(billboard100, .)
```
# Select
```{r}
billboard100 %>%
  select(date, rank, song, artist, 'weeks-on-board')

billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board')

billboard100 %>%
  select(-'last-week', -'peak-rank')
```
# Mutate
```{r}
billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board') %>%
  mutate(is_collab = grepl('Featuring', artist)) %>%
  select(artist, is_collab, everything())
```
# Filter
```{r}
billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board') %>%
  filter(weeks_popular >= 20, artist == 'Drake' | artist == 'Taylor Swift')
```

# Distinct
```{r}
billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board') %>%
  filter(artist == 'Drake') %>%
  distinct(song)

billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board') %>%
  filter(artist == 'Drake') %>%
  distinct(song) %>%
  .$song
```

# Group_by & Summarise
```{r}
billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board') %>%
  filter(artist == 'Drake') %>%
  group_by(song) %>%
  summarise(total_weeks_popular = max(weeks_popular))
```
# Arrange
```{r}
billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board') %>%
  filter(artist == 'Drake') %>%
  group_by(song) %>%
  summarise(total_weeks_popular = max(weeks_popular)) %>%
  arrange(desc(total_weeks_popular), song) %>%
  head(10)
```

# Count
```{r}
billboard100 %>%
  select(date:artist, weeks_popular='weeks-on-board')  %>%
  count(artist) %>%
  arrange(desc(n))
```

#This document includes all the examples from the warmup + HW code we have done. Please refer to the specific LAB for practice code! 
HW #1: RMarkdown (lab 1)
HW #2: Data Structures (lab 2)
HW #3. Data Frames (lab 3)
HW #4: Dplyr Practice (lab 4)
HW #5: Dplyr Superhero (lab 6.2)
HW #6: Summary Functions (HW7)

# Lab 5 Intro
```{r}
#1. Load the tidyverse 
#library(tidyverse)

#2. Load the fish data 
# fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")

#3. Transform the fish data to only include the variables lakeid and length.  Store this as a new dataframe called "fishlength".  
# names(fish) 
# fishlength <- select(fish, "lakeid", "length")

#4. Filter the `fish` data to include the samples from lake "BO".
# glimpse(fish) ;from this we find the class lakeid is chr 
# filter(fish, lakeid == "BO")

#5. Calculate the mean length of fish from lake "BO".
# fish_bo <- filter(fish, lakeid == "BO")
# mean(fish_bo$length)
```

#Lab 5 WarmUp
```{r}

```


#Lab 6 WarmUp
## Load the libraries 
```{r}
#library("tidyverse")
#library("janitor")
```

# 1. Load the data 
```{r}
#bison <- readr::read_csv("data/bison.csv")
```

# 2. Dimensions and structure of the data? 
```{r}
#dim(bison) #Dimensions
```

```{r}
#glimpse(bison) #structure of the data
```
Note: dbl is double floating point therefore its a number. 

#3. We are interested in code, sex, weight, year of birth. Restrict the data to these variables and store the dataframe as a new object.
```{r}
#bison_new <- bison %>%
  #select(animal_code, animal_sex, animal_weight, animal_yob)
```

#4. Pull out the animals born between 1980-1990  
```{r}
bison_new %>%
  filter(animal_yob>=1980 & animal_yob<=1990)
```

#5. How many male and female bison are represented betwen 1980-1990?
```{r}
males <- bison_new %>% #first pull out the males 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="M")
males #there are 21 males based on variables in the table. 

table(males$animal_sex)
```

```{r}
females <- bison_new %>% #first pull out the females 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="F")
females # 

table(females$animal_sex) #There are 414 females 
```

#6. Between 1980-1990, were males or females larger on average
```{r}
mean(males$animal_weight)
```

```{r}
mean(females$animal_weight)
```

#Lab 6.2 is the homework! 

#HW 7 Another Method of Getting Answer

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
```

For this assignment we are going to work with a large data set from the [United Nations Food and Agriculture Organization](http://www.fao.org/about/en/) on world fisheries. These data are pretty wild, so we need to do some cleaning. First, load the data.  

Load the data `FAO_1950to2012_111914.csv` as a new object titled `fisheries`.
```{r}
fisheries <- readr::read_csv(file = "data/FAO_1950to2012_111914.csv")
```

1. Do an exploratory analysis of the data (your choice). What are the names of the variables, what are the dimensions, are there any NA's, what are the classes of the variables?  
```{r}
dim(fisheries)
```

```{r}
names(fisheries)
```

```{r}
anyNA(fisheries)
```

```{r}
skim(fisheries)
```

2. Use `janitor` to rename the columns and make them easier to use. As part of this cleaning step, change `country`, `isscaap_group_number`, `asfis_species_number`, and `fao_major_fishing_area` to data class factor. 
```{r}
fisheries <- clean_names(fisheries)
```

We need to deal with the years because they are being treated as characters and start with an X. We also have the problem that the column names that are years actually represent data. We haven't discussed tidy data yet, so here is some help. You should run this ugly chunk to transform the data for the rest of the homework. It will only work if you have used janitor to rename the variables in question 2!  
```{r}
fisheries_tidy <- fisheries %>% 
pivot_longer(-c(country,common_name,isscaap_group_number,isscaap_taxonomic_group,asfis_species_number,asfis_species_name,fao_major_fishing_area,measure),
names_to = "year",
values_to = "catch",
values_drop_na = TRUE) %>% 
mutate(year= as.numeric(str_replace(year, 'x', ''))) %>% 
mutate(catch= str_replace(catch, c(' F'), replacement = '')) %>% 
mutate(catch= str_replace(catch, c('...'), replacement = '')) %>% 
mutate(catch= str_replace(catch, c('-'), replacement = '')) %>% 
mutate(catch= str_replace(catch, c('0 0'), replacement = ''))

fisheries_tidy$catch <- as.numeric(fisheries_tidy$catch)
```

3. How many countries are represented in the data? Provide a count and list their names.
```{r}
fisheries_tidy %>% 
  group_by(country) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n)) #there is 203 countries
```

4. Refocus the data only to include country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch.
```{r}
fish <- fisheries_tidy %>% 
  select(country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch)
```

5. Based on the asfis_species_number, how many distinct fish species were caught as part of these data?
```{r}
fish %>% 
  group_by(asfis_species_name, asfis_species_number) %>% 
  summarize(n=n(), .groups= 'keep') %>% 
  arrange(desc(n)) #there was 1,551 species
```

6. Which country had the largest overall catch in the year 2000?
```{r}
fish %>% 
  filter(year==2000) %>% 
  arrange(desc(catch)) #the country that had the highest catch is China
```


7. Which country caught the most sardines (_Sardina pilchardus_) between the years 1990-2000?
```{r}
fish %>% 
  filter(between(year, 1990, 2000)) %>% 
  filter(asfis_species_name=="Sardina pilchardus") %>% 
  arrange(desc(catch)) #the country that had the most sardines is Morocco
```

8. Which five countries caught the most cephalopods between 2008-2012?
```{r}
fish %>% 
  filter(between(year, 2008, 2012)) %>% 
  filter(asfis_species_name=="Cephalopoda") %>% 
  arrange(desc(catch)) #the five countries that caught the most cephalopods was India, China, Italy, Spain, and Algeria
```

9. Which species had the highest catch total between 2008-2012? (hint: Osteichthyes is not a species)
```{r}
fish %>% 
  filter(between(year, 2008, 2012)) %>% 
  filter(!asfis_species_name=="Osteichthyes") %>% 
  arrange(desc(catch)) # the species with the highest catch total was Trichiurus lepturus
```

10. Use the data to do at least one analysis of your choice.
```{r}
bh_fish <- fisheries_tidy %>% 
  select(country, common_name, year, catch) %>% 
  filter(country=="Bosnia and Herzegovina") %>% 
  filter(between(year, 1992, 1995)) 
bh_fish
```

```{r}
hist(bh_fish$year)
```