---
title: "code_MT1"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#This document includes all the examples from the warmup + HW code we have done. Please refer to the specific LAB for practice code! 
HW #1: RMarkdown (lab 1)
HW #2: Data Structures (lab 2)
HW #3. Data Frames (lab 3)
HW #4: Dplyr Practice (lab 4)
HW #5: Dplyr Superhero (lab 6.2)
HW #6: Summary Functions (? location)

# Lab 5 Intro
```{r}
#1. Load the tidyverse 
#library(tidyverse)

#2. Load the fish data 
# fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")

#3. Transform the fish data to only include the variables lakeid and length.  Store this as a new dataframe called "fishlength".  
# names(fish) 
# fishlength <- select(fish, "lakeid", "length")

#4. Filter the `fish` data to include the samples from lake "BO".
# glimpse(fish) ;from this we find the class lakeid is chr 
# filter(fish, lakeid == "BO")

#5. Calculate the mean length of fish from lake "BO".
# fish_bo <- filter(fish, lakeid == "BO")
# mean(fish_bo$length)
```

#Lab 5 WarmUp
```{r}

```


#Lab 6 WarmUp
## Load the libraries 
```{r}
#library("tidyverse")
#library("janitor")
```

# 1. Load the data 
```{r}
#bison <- readr::read_csv("data/bison.csv")
```

# 2. Dimensions and structure of the data? 
```{r}
#dim(bison) #Dimensions
```

```{r}
#glimpse(bison) #structure of the data
```
Note: dbl is double floating point therefore its a number. 

#3. We are interested in code, sex, weight, year of birth. Restrict the data to these variables and store the dataframe as a new object.
```{r}
#bison_new <- bison %>%
  #select(animal_code, animal_sex, animal_weight, animal_yob)
```

#4. Pull out the animals born between 1980-1990  
```{r}
bison_new %>%
  filter(animal_yob>=1980 & animal_yob<=1990)
```

#5. How many male and female bison are represented betwen 1980-1990?
```{r}
males <- bison_new %>% #first pull out the males 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="M")
males #there are 21 males based on variables in the table. 

table(males$animal_sex)
```

```{r}
females <- bison_new %>% #first pull out the females 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="F")
females # 

table(females$animal_sex) #There are 414 females 
```

#6. Between 1980-1990, were males or females larger on average
```{r}
mean(males$animal_weight)
```

```{r}
mean(females$animal_weight)
```

#Lab 6.2 is the homework! 
