---
title: "code_MT1"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#This document includes all the examples from the warmup + HW code we have done. Please refer to the specific LAB for practice code! 
HW #1: RMarkdown (lab 1)
HW #2: Data Structures (lab 2)
HW #3. Data Frames (lab 3)
HW #4: Dplyr Practice (lab 4)
HW #5: Dplyr Superhero (lab 6.2)
HW #6: Summary Functions (? location)

# Lab 5 Intro
```{r}
#1. Load the tidyverse 
#library(tidyverse)

#2. Load the fish data 
# fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")

#3. Transform the fish data to only include the variables lakeid and length.  Store this as a new dataframe called "fishlength".  
# names(fish) 
# fishlength <- select(fish, "lakeid", "length")

#4. Filter the `fish` data to include the samples from lake "BO".
# glimpse(fish) ;from this we find the class lakeid is chr 
# filter(fish, lakeid == "BO")

#5. Calculate the mean length of fish from lake "BO".
# fish_bo <- filter(fish, lakeid == "BO")
# mean(fish_bo$length)
```

#Lab 5 WarmUp
```{r}

```


#Lab 6 WarmUp
## Load the libraries 
```{r}
#library("tidyverse")
#library("janitor")
```

# 1. Load the data 
```{r}
#bison <- readr::read_csv("data/bison.csv")
```

# 2. Dimensions and structure of the data? 
```{r}
#dim(bison) #Dimensions
```

```{r}
#glimpse(bison) #structure of the data
```
Note: dbl is double floating point therefore its a number. 

#3. We are interested in code, sex, weight, year of birth. Restrict the data to these variables and store the dataframe as a new object.
```{r}
#bison_new <- bison %>%
  #select(animal_code, animal_sex, animal_weight, animal_yob)
```

#4. Pull out the animals born between 1980-1990  
```{r}
bison_new %>%
  filter(animal_yob>=1980 & animal_yob<=1990)
```

#5. How many male and female bison are represented betwen 1980-1990?
```{r}
males <- bison_new %>% #first pull out the males 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="M")
males #there are 21 males based on variables in the table. 

table(males$animal_sex)
```

```{r}
females <- bison_new %>% #first pull out the females 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="F")
females # 

table(females$animal_sex) #There are 414 females 
```

#6. Between 1980-1990, were males or females larger on average
```{r}
mean(males$animal_weight)
```

```{r}
mean(females$animal_weight)
```

#Lab 6.2 is the homework! 

#HW 7 Another Method of Getting Answer

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
```

For this assignment we are going to work with a large data set from the [United Nations Food and Agriculture Organization](http://www.fao.org/about/en/) on world fisheries. These data are pretty wild, so we need to do some cleaning. First, load the data.  

Load the data `FAO_1950to2012_111914.csv` as a new object titled `fisheries`.
```{r}
fisheries <- readr::read_csv(file = "data/FAO_1950to2012_111914.csv")
```

1. Do an exploratory analysis of the data (your choice). What are the names of the variables, what are the dimensions, are there any NA's, what are the classes of the variables?  
```{r}
dim(fisheries)
```

```{r}
names(fisheries)
```

```{r}
anyNA(fisheries)
```

```{r}
skim(fisheries)
```

2. Use `janitor` to rename the columns and make them easier to use. As part of this cleaning step, change `country`, `isscaap_group_number`, `asfis_species_number`, and `fao_major_fishing_area` to data class factor. 
```{r}
fisheries <- clean_names(fisheries)
```

We need to deal with the years because they are being treated as characters and start with an X. We also have the problem that the column names that are years actually represent data. We haven't discussed tidy data yet, so here is some help. You should run this ugly chunk to transform the data for the rest of the homework. It will only work if you have used janitor to rename the variables in question 2!  
```{r}
fisheries_tidy <- fisheries %>% 
pivot_longer(-c(country,common_name,isscaap_group_number,isscaap_taxonomic_group,asfis_species_number,asfis_species_name,fao_major_fishing_area,measure),
names_to = "year",
values_to = "catch",
values_drop_na = TRUE) %>% 
mutate(year= as.numeric(str_replace(year, 'x', ''))) %>% 
mutate(catch= str_replace(catch, c(' F'), replacement = '')) %>% 
mutate(catch= str_replace(catch, c('...'), replacement = '')) %>% 
mutate(catch= str_replace(catch, c('-'), replacement = '')) %>% 
mutate(catch= str_replace(catch, c('0 0'), replacement = ''))

fisheries_tidy$catch <- as.numeric(fisheries_tidy$catch)
```

3. How many countries are represented in the data? Provide a count and list their names.
```{r}
fisheries_tidy %>% 
  group_by(country) %>% 
  summarize(n=n()) %>% 
  arrange(desc(n)) #there is 203 countries
```

4. Refocus the data only to include country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch.
```{r}
fish <- fisheries_tidy %>% 
  select(country, isscaap_taxonomic_group, asfis_species_name, asfis_species_number, year, catch)
```

5. Based on the asfis_species_number, how many distinct fish species were caught as part of these data?
```{r}
fish %>% 
  group_by(asfis_species_name, asfis_species_number) %>% 
  summarize(n=n(), .groups= 'keep') %>% 
  arrange(desc(n)) #there was 1,551 species
```

6. Which country had the largest overall catch in the year 2000?
```{r}
fish %>% 
  filter(year==2000) %>% 
  arrange(desc(catch)) #the country that had the highest catch is China
```


7. Which country caught the most sardines (_Sardina pilchardus_) between the years 1990-2000?
```{r}
fish %>% 
  filter(between(year, 1990, 2000)) %>% 
  filter(asfis_species_name=="Sardina pilchardus") %>% 
  arrange(desc(catch)) #the country that had the most sardines is Morocco
```

8. Which five countries caught the most cephalopods between 2008-2012?
```{r}
fish %>% 
  filter(between(year, 2008, 2012)) %>% 
  filter(asfis_species_name=="Cephalopoda") %>% 
  arrange(desc(catch)) #the five countries that caught the most cephalopods was India, China, Italy, Spain, and Algeria
```

9. Which species had the highest catch total between 2008-2012? (hint: Osteichthyes is not a species)
```{r}
fish %>% 
  filter(between(year, 2008, 2012)) %>% 
  filter(!asfis_species_name=="Osteichthyes") %>% 
  arrange(desc(catch)) # the species with the highest catch total was Trichiurus lepturus
```

10. Use the data to do at least one analysis of your choice.
```{r}
bh_fish <- fisheries_tidy %>% 
  select(country, common_name, year, catch) %>% 
  filter(country=="Bosnia and Herzegovina") %>% 
  filter(between(year, 1992, 1995)) 
bh_fish
```

```{r}
hist(bh_fish$year)
```