---
title: "study_guide_mt1"
output: html_document
date: "2024-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Lab 1 
class repository "datascibiol" (https://github.com/jmledford3115/datascibiol)**Star**
[R cheatsheets](https://posit.co/resources/cheatsheets/)
-GitHub is a file storage and management site used by programmers. The software Git uses version control. 
--Programmers upload code to **repositories(folders)** and make it publicly avail.
-GitHub desktop is an easy way to manage GitHub repository. 
## Clone Your Repository  
In GitHub desktop, you will see your personal repository as an option under `clone repository`. Clone your repository to the desktop. Once you have done this any changes you make to files will be recorded and easily uploaded. Do not move this folder from your desktop. 
## Copy the Class Repository
You do this by navigating to the repository then going to the small green code button at top right. Click on the code button then go to Download ZIP.  
-Terms to Define: 
1. `Commit` takes a snapshot of your current progress.    
2. `Pull` this is the same as download.     
3. `Push` this is the same as upload.  
Clicking the `Commit to master` button in the bottom left corner means that you are recording a snapshot of your edits locally. Once this is done, you need to upload your changes to GitHub. This is done by clicking `Push origin`. Once you have pushed your changes you can double check by looking at your repository on github.com.    
## R and RStudio
R is an open source, **scripting** language. 
--therefore, write instructiongs used by the computer to complete an operation.
---no typos, punctuation errors, errant spaces, or other types of mistakes. 
RStudio is a GUI that is used to interact with R. 
-**working directory** is the location on your computer where you are currently working. (Place where R studio will save all your work.)
```{r}
getwd() #if not in correct wd then, Session>Set Working Directory>Choose Directory>Desktop>
```
In R '>' is where commands are entered. 
##Arithmetic
Order of operations applies. No need for = 
-Making a **vector** aka string of values to store an an object then we can perform stats on the object. 
--'Objects' in R are a way in which we can store data or operations. 
## RMarkdown Basics 
Document read in R Studio. 
**Why use R Markdown**. 
Transparency in experimental methodology, observation, collection of data and analytical methods.
Public availability and re-usability of scientific data.
Public accessibility and transparency of scientific communication.
Using web-based tools to facilitate scientific collaboration.
###Titles and Text
'#' is used for tittles 
-The more hashtags = smaller font
--Text with no '#' is smallest and not bold. 
_italics_
**bold**
line break is indiacated. 
by two spaces. 
###Links
'[]' for name of the webpage and '()' for website address. 
For emails you need to include 'malito:' in parenthesis. [Liana Williams](malito:lcwilliams@ucdavis.edu)
###Code chunks 
 "Insert>R" at the top of the page. On a Mac, the shortcut is `option+command+i`.  
###Knit 
make an html file "builds a file" 
##To get help with a command 
?**command name**

To install.packages(_insertnameofpackage_)  
once installed then, library(_nameofpackage_)
#Lab2 
Assign a value to object 'x'. The '<-' symbol is read as 'gets'. In this case, x gets 42. Make sure that you are in the environment panel and you should see the value associated with 'x'. On a mac, you can push `option` and `-` to automatically generate the gets symbol.  
##2.1 Objects, Nomenclature, Types of Data, Missing Data
-Don't assign names to objects that are the same as functions in R 
*There are five frequently used `classes` of data: 1. numeric, 2. integer(L notates an integer), 3. character(""), 4. logical(TRUE or FALSE), 5. complex.*
To find out what type of data you are working with, use the `class()` function.
You can use the `is()` and `as()` functions to clarify or specify a type of data.
```{r}
is.integer(my_numeric) #is my_numeric an integer?
```

```{r}
my_integer <- 
  as.integer(my_numeric) #create a new object specified as an integer
```

```{r}
is.integer(my_integer) #is my_numeric an integer?
```
NA values are used to designate missing data. `is.na` or `anyNA` are useful functions when dealing with NAs in data.
-Tells TRUE or FALSE 
```{r}
mean(new_vector, na.rm=T) #na.rm removes the NA values in the vector
```
##2.2 
Data Structures: include vectors, lists, matrices, dataframes, and factors 
**vectors** create vectors using the 'c' command which means concatenate. 
*A vector is a list of variables, and the simplest data structure in R*  
--We can **identify vector elements** with the [], you only get the values, not the logical evaluation of the entire vector.
**Data Matrices**: series of stacked vectors, similar to a data table. 
Name the columns using `colnames()` with the vector region.
Name the rows using `rownames()` with the vector titles.
`cbind()` adds columns.
Adding a colon `:` selects the specified elements in a column.  

????????????Question to ask in OH; in 2.2. is it read row on the left column on the right? in #a few extras I am confused. 
- if you want the entire column do something like [ ,2] which pulls out all of coumn 2 (look vertically?)


#Lab 3 
???????? For the warm up when we were making the matrix. How do you know it by nrow =5 and byrow = F? What do those functions do?
--can you review when you would do colmeans vs rowmeans???
Vectors can be organized into a new data structure called **data frames**. 
Data frames are similar to a spreadsheet. It can store data of many different classes. ?????????Can a data matrix only do one type of class?  
##3.1 
Ex: Combine our three vectors to create a data frame with the function `data.frame()`.  
Data Frames show the class of data.
'names()' gives the column names of the vector  
'dim()' gives the dimensions
'str()' gives the structure; this gives the structure where each variable(column) is NOW a row, an observation is what goes in a cell.
`dbl` means that the value is a type of numeric [double precision floating point](http://uc-r.github.io/integer_double/).  
We can change the name in a dataframe by (new_name=Old_name) the data in the dataframe is overwritten so it now appears as new_name for the variable. 
## Accessing Data Frame Columns and Rows 
The same methods of selecting elements in vectors and data matrices apply to data frames. We use `[]`. We have two positions where the first applies to the rows, and the second to the columns.  
Ex: hbirds[1,] means first row; if [,3] then, 3rd column.
We can also select values in an entire column using the `$` sign.  
We can add a new bird's data with the `rbind()` function, known as row bind. It will bind our new vector to our data frame row-wise.
## Writing Data to File  
.csv is comma separated values (data type that can be imported into R)

##3.2 
1. Import .csv files as data frames using `read_csv()`.  
##Load the Data 
Since we are using the tidyverse, we open the file using `read_csv()`. `readr` is included in the tidyverse set of packages. We specify the package and function with the `::` symbol. This becomes important if you have multiple packages loaded that contain functions with the same name. 
A **factor** has discrete levels (there is a max. amount). 
## Summary functions
We can summarize our data frame with the`summary()` function.  
`glimpse()` is another useful summary function.  
`nrow()` gives the numbers of rows.  
`ncol` gives the number of columns.  
`names` gives the column names.  
`head()` prints the first n rows of the data frame.  
`tail()` prinst the last n rows of the data frame.  
`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable. We will come back to this later...   
We can also 'View()' in the environment panel  
## Filter
Filter is a way of pulling out observations that meet specific criteria in a variable. (More in lab 4).   
Observations are in a cell of the spreadsheet  

