---
title: "Lab 4 Homework"
author: "Liana Williams"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Data
For the homework, we will use data about vertebrate home range sizes. The data are in the class folder, but the reference is below.  

**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

**1. Load the data into a new object called `homerange`.**
###Load the data 
```{r}
homerange <- readr::read_csv("data/Tamburelloetal_HomeRangeDatabase.csv")
#readr is explicit of the package we are use
```

**2. Explore the data. Show the dimensions, column names, classes for each variable, and a statistical summary. Keep these as separate code chunks.**  
```{r}
dim(homerange)
#This code shows the dimensions of the homerange data 
```
```{r}
names(homerange)
#This codes shows the column names 
```

```{r}
str(homerange)
#This shows the classes for each variable 
```

```{r}
summary(homerange)
#This shows a statistical summary
```

**3. Change the class of the variables `taxon` and `order` to factors and display their levels.**  
```{r}
homerange$taxon <- as.factor(homerange$taxon)
levels(homerange$taxon) 
#replace taxon in the homerange dataframe but make it into a factor. Then display the levels
```

```{r}
homerange$order <- as.factor(homerange$order)
levels(homerange$order)
```

**4. What taxa are represented in the `homerange` data frame? Make a new data frame `taxa` that is restricted to taxon, common name, class, order, family, genus, species.**  
```{r}
taxa <- select(homerange, "taxon", "common.name", "class", "order", "family", "genus", "species")
taxa
```

**5. The variable `taxon` identifies the common name groups of the species represented in `homerange`. Make a table the shows the counts for each of these `taxon`.**  
```{r}
table(homerange$taxon)
```

**6. The species in `homerange` are also classified into trophic guilds. How many species are represented in each trophic guild.**  
```{r}
table(homerange$trophic.guild)
```

**7. Make two new data frames, one which is restricted to carnivores and another that is restricted to herbivores.**  
```{r}
carnivores <- filter(homerange, trophic.guild == "carnivore")
herbivores <- filter(homerange, trophic.guild == "herbivore")
```

**8. Do herbivores or carnivores have, on average, a larger `mean.hra.m2`? Remove any NAs from the data.**  
```{r}
mean(herbivores$mean.hra.m2, rm.na=T)
```

```{r}
mean(carnivores$mean.hra.m2, rm.na=T)
``` 

**9. Make a new dataframe `owls` that is limited to the mean mass, log10 mass, family, genus, and species of owls in the database. Which is the smallest owl? What is its common name? Do a little bit of searching online to see what you can learn about this species and provide a link below** 
```{r}
owls <- filter(homerange, order=="strigiformes")
owls_df <- select(owls, "mean.mass.g", "log10.mass", "family", "genus", "species")
filter(owls_df, mean.mass.g==61.32) #This finds the specific small owl
```
Answer  
The smallest owl appears to be the Glaucidium Passerinum, the Eurasian pygmy owl (is the common name).  
I looked it up and found https://en.wikipedia.org/wiki/Eurasian_pygmy_owl from this the species has spots on its sides and is found in the "boreal forests of Northern and Central Europe to Siberia."

```{r}
#owls_datf <- select(filter(homerange, order=="strigiformes"), mean.mass.g, log10.preymass, family, genus, species)
#owls_datf
#all birds belongs to aves 
#all owls belong to one order called strigiformes
#To get the common name I added it to the comma list as common.name,
```
#^This was another methods that I found to work

**10. As measured by the data, which bird species has the largest homerange? Show all of your work, please. Look this species up online and tell me about it!**.  
```{r}
birds <- filter(homerange, taxon=="birds")
max(birds$mean.hra.m2)
```
```{r}
filter(birds, mean.hra.m2==2.41e+08)
```
Answer   
The bird species with the largest homerange is caracara cheriway.  
https://animaldiversity.org/accounts/Caracara_cheriway/  
The Caracara cheriway is present along the Mexican American Border from Baja California to Eastern Texas, then south to Panama. Their habitats are comprised of dry prairie with some wetter areas. Cararas have long yellow legs, and a large, hooked, bluish bill.   

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.   