---
title: "lab6_warmup"
output: html_document
date: "2024-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the libraries 
```{r}
library("tidyverse")
library("janitor")
```

# 1. Load the data 
```{r}
bison <- readr::read_csv("data/bison.csv")
```
# 2. Dimensions and structure of the data? 
```{r}
dim(bison) #Dimensions
```

```{r}
glimpse(bison) #structure of the data
```
Note: dbl is double floating point therefore its a number. 

#3. We are interested in code, sex, weight, year of birth. Restrict the data to these variables and store the dataframe as a new object. 
```{r}
#bison_new <- select(bison, "animal_code", "animal_sex", "animal_weight", "animal_yob")
#this way works but, prof wants to use pipes
```

```{r}
bison_new <- bison %>%
  select(animal_code, animal_sex, animal_weight, animal_yob)
```

#4. Pull out the animals born between 1980-1990 
```{r}
#filter(bison, between(animal_yob, 1980, 1990)) #pull all observations for animal_yob 1980-1990
#this way works 
```
ˆ
```{r}
bison_new %>%
  filter(animal_yob>=1980 & animal_yob<=1990)
```

#5. How many male and female bison are represented betwen 1980-1990?
```{r}
males <- bison_new %>% #first pull out the males 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="M")
males #there are 21 males based on variables in the table. 

table(males$animal_sex)
```
```{r}
females <- bison_new %>% #first pull out the females 
  filter(animal_yob>=1980 & animal_yob<=1990) %>%
  filter(animal_sex=="F")
females # 

table(females$animal_sex) #There are 414 females 
```


#6. Between 1980-1990, were males or females larger on average
```{r}
mean(males$animal_weight)
```

```{r}
mean(females$animal_weight)
```

