---
title: "summarize practice, `count()`, `across()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Produce clear, concise summaries using a variety of functions in `dplyr` and `janitor.`  
2. Use the `across()` operator to produce summaries across multiple variables.  


####FIX THE CODE SO IT MATCHES THE PROFS###

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("skimr")
library("palmerpenguins")
```

## Review
The summarize() and group_by() functions are powerful tools that we can use to produce clean summaries of data. Especially when used together, we can quickly group variables of interest and save time. Let's do some practice with the [palmerpenguins(https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data to refresh our memory.
```{r}
glimpse(penguins)
```

As biologists, a good question that we may ask is how do the measured variables differ by island (on average)?
```{r}
levels(penguins$island) #This tells us there are three islands. island is a factor so use levels. 
```

```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(mean_body_mass=mean(body_mass_g))
```

**On which island is body mass highest** (Potential midterm question)
**group_by only works on categorical data** 


Why do we have NA here? Do all of these penguins lack data?
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(number_NAs=sum(is.na(body_mass_g))) #The NAs that exist are messing up the data 
```

Well, that won't work so let's remove the NAs and recalculate.
```{r}
penguins %>% 
  filter(!is.na(body_mass_g)) %>% #pull out all the observations that have a number and leave all the NAs behind. 
  
```

What if we are interested in the number of observations (penguins) by species and island?
```{r}
penguins %>% 
  group_by(species, island) %>% 
  summarize(n=n(), .groups= 'keep')#the .groups argument here just prevents a warning message
#the n=n() is the number of observations! 
```

**This is important to the Exam (counts)**

## Counts
Although these summary functions are super helpful, oftentimes we are mostly interested in counts. The [janitor package](https://garthtarr.github.io/meatR/janitor.html) does a lot with counts, but there are also functions that are part of dplyr that are useful.  

`count()` is an easy way of determining how many observations you have within a column. It acts like a combination of `group_by()` and `n()`.
```{r}
penguins %>% 
  count(island, sort = T) #sort=T sorts the column in descending order; group by the island and gives the number easily
```
**These two are the same^**

Compare this with `summarize()` and `group_by()`.
```{r}
penguins %>% 
  group_by(island) %>% 
  summarize(n=n())
```

You can also use `count()` across multiple variables.
```{r}
penguins %>% 
  count(island, species, sort = T) # sort=T will arrange in descending order; multiple variables in this example are island and species. 
```

For counts, I also like `tabyl()`. Lots of options are supported in [tabyl](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html)
```{r}
penguins %>% 
  tabyl(island, species) #tabyl is apart of janitor
```

## Practice
1. How does the mean of `bill_length_mm` compare between penguin species?
```{r}
penguins %>% 
  group_by(species) %>% 
  summarize(mean_bill_length_mm=mean(bill_length_mm, na.rm = T))
```

2. For some penguins, their sex is listed as NA. Where do these penguins occur?
```{r}
penguins %>% 
  count(sex, island)
```
some observations on each island where it hasn't been recorder. there could have been some problem recording the data. 

## `across()`
What about using `filter()` and `select()` across multiple variables? There is a function in dplyr called `across()` which is designed to work across multiple variables. Have a look at Rebecca Barter's awesome blog [here](http://www.rebeccabarter.com/blog/2020-07-09-across/).    

What if we wanted to apply `summarize()` in order to produce distinct counts over multiple variables; i.e. species, island, and sex? Although this isn't a lot of coding you can image that with a lot of variables it would be cumbersome.
```{r}
penguins %>%
  summarize(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_sex = n_distinct(sex))
```

By using `across()` we can reduce the clutter and make things cleaner. 
```{r}
penguins %>%
  summarize(across(c(species, island, sex), n_distinct)) #variables of interest, ndistinct
```

This is very helpful for continuous variables.
```{r}
penguins %>%
  summarize(across(contains("mm"), mean, na.rm=T))
#summarize across all the variables that mm 
```

```{r}
penguins %>%
  summarize(across(contains("mm"), \(x) mean(x, na.rm = TRUE))) #use this to correct the error
```
------------The examples after this are not going to make or break you on the midterm--------------------e
`group_by` also works.
```{r}
penguins %>%
  group_by(sex) %>% 
  summarize(across(contains("mm"), mean, na.rm=T))
```

Here we summarize across all variables.
```{r}
penguins %>%
  summarise_all(mean, na.rm=T)
```

Operators can also work, here I am summarizing across all variables except `species`, `island`, `sex`, and `year`.
```{r}
penguins %>%
  summarise(across(!c(species, island, sex, year), 
                   mean, na.rm=T))
```

All variables that include "bill"...all of the other dplyr operators also work.
```{r}
penguins %>%
  summarise(across(starts_with("bill"), mean, na.rm=T))
```

## Practice
1. Produce separate summaries of the mean and standard deviation for bill_length_mm, bill_depth_mm, and flipper_length_mm for each penguin species. Be sure to provide the number of samples.  
```{r}

```

## Wrap-up  

Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)